---
title: "Coursera Reproducible Research Project 1"
author: "Pachawo Bisani"
date: "9/21/2019"
output: html_document
---
# Set up environment.

```{r setup}
library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(VIM)
library(mice)
set.seed(1010)
```

## 1. Read csv data

The code below is used to read activity.csv data:

```{r pressure, echo=TRUE}
activity <- read_csv("activity.csv")
```

A function read_csv from readr package has been used. The choice of read_csv over read is because read_csv handles dates automatically unlike read function that requires dates values be coerced from class character to date.

## 2. Histogram of the total number of steps taken each day

```{r echo=TRUE}
activity %>% group_by(date) %>% summarise(stepsPerDay = sum(steps)) %>% 
  ggplot(aes(x = stepsPerDay)) + geom_histogram(bins = 15) + 
  ggtitle("Histogram: Total Number of Steps Per Day") + 
  xlab("Steps Per Day") + ylab("Frequency")
```
## 3. Mean and median number of steps taken each day

### Mean steps, missing values removed
```{r echo=TRUE}
activity %>% group_by(date) %>% 
  summarise(meanStepsPerDay = mean(steps, na.rm = TRUE)) %>% 
  summarise(meanSteps = mean(meanStepsPerDay, na.rm = TRUE))
```

### Median steps, missing values removed
```{r echo=TRUE}
activity %>% group_by(date) %>% 
  summarise(medianStepsPerDay = median(steps, na.rm = TRUE)) %>%
  summarise(medianSteps = median(medianStepsPerDay, na.rm = TRUE))
```

## 4.Time series plot of the average number of steps taken

```{r echo=TRUE}
activity %>% group_by(date) %>% summarise(meanSteps = mean(steps, na.rm = TRUE)) %>%
  ggplot(aes(x = date, y = meanSteps)) + geom_line() + 
  ggtitle("Mean Steps by Date") + xlab("Date") + ylab("Mean Steps")
```

## 5. The 5-minute interval that, on average, contains the maximum number of steps
```{r echo=TRUE}
activity %>% group_by(interval) %>% 
  summarize(meanByInterval = mean(steps, na.rm = TRUE)) %>%
  filter(meanByInterval == max(meanByInterval))
```

## The 5-minute interval that, on average, contains the minimum number of steps

```{r echo=TRUE}
activity %>% group_by(interval) %>% 
  summarize(meanByInterval = mean(steps, na.rm = TRUE)) %>%
  filter(meanByInterval == min(meanByInterval))
```

## 6. Code to describe and show a strategy for imputing missing data

### Missing values by variable

```{r echo=TRUE}
md.pattern(activity)
```

```{r echo=TRUE}
(missing <- sum(is.na(activity)))
```

### Missing Values as percent of total, percent each column

```{r echo=TRUE}
missingPercent <- sum(is.na(activity))/(dim(activity)[1]*dim(activity)[2]) * 100 
pMiss <- function(x) { sum(is.na(x)) / length(x) * 100}
(missingPercentCol <-apply(activity, 2, pMiss))
```

### Missing values visualization

```{r echo=TRUE}
aggr(activity, numbers = TRUE)
```

### Missing values strategy: take complete cases only

```{r echo=TRUE}
activityNoMissing <- activity[complete.cases(activity),]
```

## 7. Histogram of the total number of steps taken each day after missing values are removed

```{r echo=TRUE}
activityNoMissing %>% group_by(date) %>% summarise(stepsPerDay = sum(steps)) %>% 
  ggplot(aes(x = stepsPerDay)) + geom_histogram(bins = 15) + 
  ggtitle("Histogram: Total Number of Steps Per Day") + xlab("Steps Per Day") + 
  ylab("Frequency")
```

## 7a. Time series plot of average number of steps taken after missing values are removed

```{r echo=TRUE}
activityNoMissing %>% group_by(date) %>% summarise(meanSteps = mean(steps, na.rm = TRUE)) %>%
  ggplot(aes(x = date, y = meanSteps)) + geom_line() + 
  ggtitle("Mean Steps by Date") + xlab("Date") + ylab("Mean Steps")
```

## 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

### By Weekday vs. Weekend
```{r echo=TRUE}
t <- activityNoMissing %>% mutate(dayOfWeek = weekdays(date)) %>%
  mutate(Weekend = ifelse(dayOfWeek == "Saturday" | dayOfWeek == "Sunday", "Weekend", "Weekday"))
```

```{r echo=TRUE}
## By Weekday vs. Weekend 
t %>% 
  group_by(Weekend, interval) %>% mutate(meanStepsInterval = mean(steps)) %>%
  ggplot(aes(x = interval, y = meanStepsInterval)) + geom_line() +
  facet_wrap(~Weekend) +ggtitle("Mean Steps by Interval: Weekday vs. Weekend") + 
  xlab("Interval") + ylab("Mean Steps")
```

### By Days of the Week
```{r echo=TRUE}
t %>%
  group_by(dayOfWeek, interval) %>% mutate(meanStepsInterval = mean(steps)) %>%
  ggplot(aes(x = interval, y = meanStepsInterval)) + geom_line() +
  facet_wrap(~dayOfWeek) +ggtitle("Mean Steps by Interval: By Day") +
  xlab("Interval") + ylab("Mean Steps")
```